Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-11-25 proposal/proposal.tex
Working out final additions. Use xelatex so section refs fixed.
There seems to be a problem with section label references using the
-pdf command of latexmk requiring a second compile to fixed it.
-xelatex does not seem suffer from this deficiency.

I have plans for discussing a Caires and Pfenning-style example.
I have plans for presenting cofinite quantification versions of
some of Wadlers typing rules.
I have plans to have a couple more plans.

% Include Coq lstlisting language definition and things upon which it depends.
\usepackage{macros}

\usepackage{hyperref}
\usepackage{listings}
\usepackage{lstcoq}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{cmll}
\usepackage[style=alphabetic,natbib=true,maxbibnames=100]{biblatex}
\addbibresource{proposal.bib}

@@

\newcommand{\fpop}{System F${}^\circ$\xspace}

@@

From this framework, a number of directions can be explored including
session polymorphism, session subtyping, and permitting different forms of
aliasing within the type system. The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be a interesting place to start when looking to add these extensions.

Additionally, the majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems. One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}. We could look to extend the GV type system to provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}. The ``adoption and focus'' type system is non-trivial requiring special language features, e.g. to manage capabilities which can be thought as a form of object liveness mechanism, so the GV typing judgements must be extended such that the capabilities can be included in soundness results. A translation to CP must also be defined for the ``adopt'' and ``focus'' constructs.

@@

\subsubsection{In the $\pi$-calculus}\label{sec:pis}

\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection between $\pi$-calculus type system with session types and intuitionistic linear logic. They equate linear propositions with session types of the process calculus via a bidirectional translation relation. They use a labelled transition system to represent process reduction with observable actions on channels. They represent dual intuitionistic linear logic as a sequent calculus providing logical connectives which have session-based interpretations. Process reduction steps are equated with proof reduction via cut elimination. The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication for the process calculus. The correspondence of session types to linear logic propositions is an important step in providing a solid logical foundation for a concurrent type theory; in the same way ``sequential'' type theory ($\lambda$-calculi) profitted from a logical perspective. However, for session types to become a more usable tool for programming distributed communication systems (the ultimate goal, of course, is to write more correct programs), we cannot be expected to program directly in the $\pi$-calculus; a more mainstream programming paradigm must be built upon these solid logical foundations. With that in mind, enter the $\lambda$-calculus!

\subsubsection{In the $\lambda$-calculus}\label{sec:lam}

@@

For handling binder freshness, I chose to adopt the cofinite quantification approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which excludes a finite set of variables from being considered as the binder. In contrast, the traditional ``exists-fresh'' approach, where the binder is only required to be fresh within the abstraction's body, does not produce strong enough induction principles in some cases.

@@

>>>>

% Include Coq lstlisting language definition and things upon which it depends.
\usepackage{macros}
\usepackage{hyperref}

\usepackage{listings}
\usepackage{lstcoq}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{cmll}
\usepackage[style=alphabetic,natbib=true,maxbibnames=100]{biblatex}
\addbibresource{proposal.bib}

@@

\newcommand{\fpop}{System F${}^\circ$\xspace}
\newcommand{\1}{\textbf{1}\xspace}

@@

From this framework, a number of directions can be explored including
session polymorphism, session subtyping, and permitting different forms of
aliasing within the type system. The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be an interesting place to start when looking to add these extensions.

Additionally, the majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems. One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}. The ``adoption and focus'' type system is non-trivial requiring special language features, e.g. to manage capabilities which can be thought as a form of object liveness mechanism, so the GV typing judgements must be extended such that the capabilities can be included in soundness results. A translation to CP must also be defined for the ``adopt'' and ``focus'' constructs.

@@

\subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}

\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection between a $\pi$-calculus type system with session types ($\pi$DILL) and dual intuitionistic linear logic (DILL)~\cite{Barber:1996}. They equate linear propositions with session types of the process calculus via a bidirectional translation relation. They use a binary reduction relation and a labelled transition system to represent process reduction and process action with its environment as is standard~\cite{Sangiorgi:2001}.

They represent dual intuitionistic linear logic as a sequent calculus providing logical connectives which have session-based interpretations in $\pi$DILL. Consider their rules for branch and choice in $\pi$DILL

... here are the rules ...

Note that branch and choice both have some notion of branching (R rules for +/L rules for \&) and at least for \& this is counter-intuitive in session-based setting, since the intutition for \& is that of offering a choice between alternatives rather than making the choice. Here, however, the intuition should be extended to account for its location in the typing judgement. The channel is on the left of the sequent which caires and pfenning suggest as some service providing to the process.

While in the case of branch and choice, which have different semantics this is mitigated somewhat but as can be seen, there are two rules for selecting a branch, one for branch and another for choice if we consider their server example

As noted by Wadler...jfp there is a certain amount of redundancy in the two-sided nature of the rules forcing one to commit to a certain convention when typing a communication system. Consider, an equivalent system to the buy/quote example given by caires and pfen. In the following, the types Book, Quote and Receipt can be considered synoms for the type \1 as in caires and pfenning.

... server example here ...

The derivation for ClientBody is T1R, T implies R, T tensor R, T plus R2. Except for the final rule, these are exact opposites of the rules used for ...

The final system is near identical to the one produced by the example given caires and pfenning. Here the client communicates along the channel $c$ on the right and the server utilises the channel $c$ on the left. This typing, while correct, violates the intuition given by caires and pfenning of the context providing the services used by the process. The example above suggests the server utilises some channel provided by the client. On the other hand, one could imagine such a channel being setup between two communicating agents; the ``owner'' of the channel is largely irrelevant. Note that clientproto is in fact the dual of serverproto in Wadler's classical process calculus.

then it is clear the server is providing a choice to the client indicated by the branch (\&). Yet an equal derivation can be obtained

Consider a shopper which operates according to a certain protocol. the idea being that the shopper can conduct transactions with any server which provides the desired service. Concretely, suppose the shopper operates according to the following protocol (a similar example to the one produced by caires and pfenning ... and inspired by gay and vasco)

choice (buy : ! Book . ! Card . ? Receipt . end, quote : ! Book . ? Price . end)

then in $\pi$DILL we can construct a shopper protocol (treating book, card, address

Process reduction steps are equated with proof reduction via cut elimination. The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication for the process calculus.

The correspondence of session types to linear logic propositions is an important step in providing a solid logical foundation for a concurrent type theory; in the same way ``sequential'' type theory ($\lambda$-calculi) profitted from a logical perspective. However, for session types to become a more usable tool for programming distributed communication systems (the ultimate goal, of course, is to write more correct programs), we cannot be expected to program directly in the $\pi$-calculus; a more mainstream programming paradigm must be built upon these solid logical foundations. With that in mind, enter the $\lambda$-calculus!

\subsubsection{In the \texorpdfstring{$\lambda$}{lambda}-calculus}
\label{sec:lam}

@@

For handling binder freshness, I chose to adopt the cofinite quantification approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which excludes a finite set of variables from being considered as the binder. In contrast, the traditional ``exists-fresh'' approach, where the binder is only required to be fresh within the abstraction's body, does not produce strong enough induction principles in some cases. This change will necessitate changes to the typing rules of GV and CP. Consider for instance the connect with rule in GV~\cite{Wadler:2012}:

... connect with rule here ... ... connect rule using cofinite quantification

The conclusion binds a variable

Consider also the rule for output in CP~\cite{Wadler:2012} and its corresponding version used here:

... output in CP ... ... version used here with cofinite quantification ...

The rule here ... is this needed? What are we showing here with the different version?

@@

 For this step of the work, I allocated one month. For the remainder, I shall look to extend the definitions of GV and CP using the definitions I have provided
